---
title: "Cognitive Skill Formation in Brazil, China, and Russia"
subtitle: "An Education Production Function Approach"
author: "Garen Avanesian"
institute: "Southern Federal University, Rostov-on-Don, Russia"
date: ""
format:
  revealjs:
    theme: [simple]
    slide-number: true
    fig-numbering: false
    chalkboard: true
    preview-links: auto
    footer: "CHLR Symposium 2025"
    incremental: false
    transition: slide
    background-transition: fade
    highlight-style: github
    scrollable: true
    controls: true
    progress: true
    embed-resources: false
    self-contained: false
    title-slide-attributes:
      data-background-color: "#2c5aa0"
editor: visual
csl: ../04_manuscript/apa-with-abstract.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r load-data}
#| echo: false
#| message: false
#| warning: false
#| output: false  
#| results: 'hide'

# Set working directory and paths for the analysis repo
projectFolder <- "/Users/karavan88/Documents/GitHub/EduProdFunctionPISA"

# Change to the analysis repo directory temporarily
setwd(projectFolder)

# Now source the project config from within the repo
source("project_config.R")

# Source the analysis scripts (suppress all output)
invisible(capture.output({
    source(file.path(rCodes, "02_descriptive.R"))
    source(file.path(rCodes, "03_modeling.R"))
}))
```

## Research Motivation {.center}

::::: columns
::: {.column width="50%"}
**The Challenge**

-   Traditional education metrics focus on *years* of schooling
-   Overlook *quality* of learning and skill formation
-   Growing inequality in educational outcomes
:::

::: {.column width="50%"}
**SDG 4.1.1c**

*Ensure all students acquire knowledge and skills for sustainable development*

-   Reading (literacy)
-   Mathematical competencies
:::
:::::

::: notes
We're addressing a fundamental question in education economics: not just how long students stay in school, but what they actually learn during that time.
:::

------------------------------------------------------------------------

## Context & Data {.smaller}

::::: columns
::: {.column width="60%"}
**PISA 2018 Assessment**

-   15-year-old students
-   Reading, Math, Science
-   Student & school questionnaires

**Sample Coverage**

-   **Brazil**: National sample
-   **China**: B-S-J-Z, Hong Kong, Macao
-   **Russia**: National + Moscow, Moscow Region, Tatarstan
:::

::: {.column width="40%"}
**Why These Countries?**

✓ Similar economic tier (upper-middle income)

✓ High inequality in learning outcomes

✓ Vastly different performance

✓ Policy relevance
:::
:::::

------------------------------------------------------------------------

## Research Questions {.smaller}

::: {.callout-important icon="false"}
## Primary Question

**What are the principal factors contributing to learning gains among secondary school students in Brazil, China, and Russia?**
:::

::: {.callout-note icon="false"}
## Secondary Question

**Are learning gains heterogeneous with respect to:**

-   Students' socio-economic status?
-   Socio-economic composition of schools?
:::

::: fragment
Focus on **net productivity** (learning gains per year) vs. **gross productivity** (total achievement levels)
:::


## The Puzzle: Gross vs. Net Productivity

```{r}
#| label: fig-pres-gross
#| fig-cap: "Average Reading Scores by Country/Territory"
#| fig-width: 10
#| fig-height: 5

library(ggplot2)

# Create a summary of average scores
gross_prod <- 
  descr_cnt %>%
  select(`Country/Territory`, Total) %>%
mutate(Territory = `Country/Territory`)

ggplot(gross_prod, aes(x = Territory, y = Total, fill = Territory)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  coord_flip() +
  scale_fill_viridis_d(option = "turbo") +
  labs(x = NULL, 
       y = "Reading Score (SD from sample mean)",
       title = "Gross Educational Productivity",
       subtitle = "Average achievement levels vary dramatically") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 16),
        panel.grid.major.y = element_blank())
```

------------------------------------------------------------------------

## Methodology: Education Production Function

::: incremental
-   **Value-added approach**: Grade progression as proxy for annual learning
-   **Mixed-effects model**: Random intercepts for schools and age groups
-   **Controls**: Individual, household, and school-level factors
-   **Separate models** for each territory (8 samples)
:::

:::: fragment
::: callout-tip
## Key Innovation

Random effects control for endogeneity from age-grade variation and school clustering
:::
::::

------------------------------------------------------------------------

## Learning Gains Vary! {.smaller}

```{r}
#| label: fig-pres-learning-gains
#| fig-cap: "Annual Learning Gains Across Territories"
#| fig-width: 10
#| fig-height: 5.5

# Extract grade coefficients from models
learning_gains <- data.frame(
  Territory = c("Hong Kong", "Tatarstan", "Russia", "BSJZ", 
                "Brazil", "Macao", "Moscow City", "Moscow Region"),
  Gain = c(0.202, 0.240, 0.264, 0.326, 0.325, 0.328, 0.347, 0.376),
  SE = c(0.028, 0.033, 0.027, 0.018, 0.013, 0.017, 0.036, 0.054)
)

learning_gains <- learning_gains %>%
  mutate(Territory = factor(Territory, levels = Territory[order(Gain)]),
         Lower = Gain - 1.96*SE,
         Upper = Gain + 1.96*SE)

ggplot(learning_gains, aes(x = Territory, y = Gain, color = Territory)) +
  geom_pointrange(aes(ymin = Lower, ymax = Upper), size = 1, linewidth = 1.2) +
  coord_flip() +
  scale_color_viridis_d(option = "turbo") +
  labs(x = NULL,
       y = "Learning Gain per School Year (SD)",
       title = "Net Educational Productivity: Learning Gains",
       subtitle = "Moscow Region shows highest annual gains despite moderate overall levels") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 16),
        panel.grid.major.y = element_blank())
```

::: fragment
**Paradox**: High gross productivity ≠ high net productivity (e.g., Hong Kong)
:::

------------------------------------------------------------------------

## Schools Matter, But How Much? {.smaller}

```{r}
#| label: fig-pres-icc
#| fig-cap: "Variance Explained by Schools (ICC)"
#| fig-width: 10
#| fig-height: 5

print(icc_plot)
```

::: incremental
-   **High stratification**: BSJZ & Brazil (48-51% school variance)
-   **Moderate**: Hong Kong, Macao, Russia (30-35%)
-   **Low stratification**: Moscow City (14%) - quality is uniform
:::

------------------------------------------------------------------------

## The Peer Effect Dominates {.smaller}

:::::: columns
::: {.column width="50%"}
**School SES Composition**

One SD increase in average school SES equals:

-   🇭🇰 **3.5 years** of schooling (Hong Kong)
-   🇷🇺 **2.5 years** (Moscow City)
-   🇷🇺 **2.0 years** (Russia, Tatarstan)\
-   🇧🇷 **0.95 years** (Brazil)
:::

:::: {.column width="50%"}
::: callout-warning
## Policy Implication

Even with same curriculum, teachers, and resources, students in high-SES schools learn substantially more

→ **Peer composition drives inequality**
:::
::::
::::::

::: fragment
The "Matthew Effect" in education: advantage begets further advantage
:::

------------------------------------------------------------------------

## Non-Cognitive Skills Matter {.smaller}

```{r}
#| label: fig-pres-eyos-task
#| fig-cap: "Effect of Task Mastery in Equivalent Years of Schooling"
#| fig-width: 10
#| fig-height: 5

# Extract task mastery effects and convert to EYOS
task_effects <- data.frame(
  Territory = c("BSJZ", "Macao", "Hong Kong", "Moscow City", 
                "Tatarstan", "Moscow Region", "Russia", "Brazil"),
  TaskCoef = c(0.033, 0.034, 0.034, 0.058, 0.044, 0.070, 0.059, 0.073),
  GradeCoef = c(0.326, 0.328, 0.202, 0.347, 0.240, 0.376, 0.264, 0.325)
)

task_effects <- task_effects %>%
  mutate(EYOS = TaskCoef / GradeCoef,
         Territory = factor(Territory, levels = Territory[order(EYOS)]))

ggplot(task_effects, aes(x = Territory, y = EYOS, fill = Territory)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_d(option = "turbo") +
  labs(x = NULL,
       y = "Equivalent Years of Schooling",
       title = "Impact of Task Mastery (1 SD increase)",
       subtitle = "Non-cognitive skills boost learning: 0.10 to 0.22 years per SD") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 16),
        panel.grid.major.y = element_blank())
```

::: fragment
"**Skills beget skills**" - malleable trait with high returns for disadvantaged students
:::

------------------------------------------------------------------------

## Gender & Language Gaps Persist {.smaller}

::::: columns
::: {.column width="50%"}
**Boys lag behind girls**

-   Tatarstan: -0.86 EYOS
-   Moscow Region: -0.66 EYOS
-   Russia, Hong Kong: -0.62 EYOS
-   Significant across all territories (except Macao)
:::

::: {.column width="50%"}
**Language minorities disadvantaged**

-   Macao, BSJZ, Moscow: \>1 year loss
-   Russia, Brazil, Tatarstan: 0.5-0.7 years loss
-   Not significant only in Hong Kong
:::
:::::

:::: fragment
::: callout-note
These are **exogenous factors** - require targeted interventions, not general policy levers
:::
::::

------------------------------------------------------------------------

## School Resources Show Mixed Results {.smaller}

::::: columns
::: {.column width="50%"}
**Significant Effects**

✓ **Teacher qualification** (Brazil, Tatarstan)\
- Brazil: 2.3 EYOS per 1% increase - Tatarstan: 0.5 EYOS

✓ **Urban location** (Russia, Tatarstan)\
- 0.6-1.0 EYOS advantage

✓ **Student-teacher ratio** (Russia, Brazil)\
- Negative effects when overcrowded
:::

::: {.column width="50%"}
**Insignificant or Unclear**

✗ School infrastructure (self-reported)

✗ Teacher qualification in high-SES areas (ceiling effect?)

✗ Class size (curvilinear relationship)

✗ Private schools (only Brazil shows effect)
:::
:::::

------------------------------------------------------------------------

## SES Shapes Learning Differently {.smaller}

```{r}
#| label: fig-pres-ses-gains
#| fig-cap: "Learning Gains by Family SES"
#| fig-width: 10
#| fig-height: 5.5

print(bar_plot_m2)
```

::: incremental
-   **Brazil**: Wealthier students learn MORE per year (privilege compounds)
-   **Russia, BSJZ**: Poorer students learn MORE per year (catching up, but from lower base)
-   **Hong Kong, Macao**: No difference in learning rates (equity in gains)
:::

------------------------------------------------------------------------

## Key Takeaways {.smaller}

::: incremental
1.  **Net ≠ Gross**: High achievement doesn't mean high learning gains
    -   Policy should target efficiency, not just outcomes
2.  **Peer composition is destiny**: Schools replicate inequality
    -   2-3 years of schooling advantage from attending high-SES schools
    -   Desegregation policies critical
3.  **Institutional stratification varies widely**
    -   China: Highly stratified (51% school variance)
    -   Russia: Moderate stratification, but huge regional gaps
    -   Policy focus: Reduce between-school inequality
4.  **Non-cognitive skills offer a lever**
    -   Malleable, high returns, especially for disadvantaged
    -   Cost-effective intervention target
:::

------------------------------------------------------------------------

## Policy Implications {.smaller}

::::: columns
::: {.column width="50%"}
**Supply-Side (Schools)**

-   Reduce school segregation by SES
-   Invest in teacher quality (context-dependent)
-   Address urban-rural gaps
-   Focus on Moscow model: high quality + low variance
:::

::: {.column width="50%"}
**Demand-Side (Students)**

-   Develop task mastery & achievement motivation
-   Support language minorities
-   Address gender gaps in reading
-   Target interventions at bottom 40% SES
:::
:::::

:::: fragment
::: callout-important
## Central Insight

Educational inequality is **jointly driven** by institutional structure and peer composition - neither alone explains outcomes
:::
::::

------------------------------------------------------------------------

## Limitations & Future Research {.smaller}

**Data Constraints**

-   Cross-sectional (not true panel data)
-   China: Urban samples only (no rural comparison)
-   Missing infrastructure data

**Methodological**

-   Peer effects: Average measures (not actual social networks)
-   Causal inference limited despite random effects
-   Omitted variable bias concerns

::: fragment
**Next Steps**

→ Panel data for true value-added estimation\
→ Full China coverage (urban-rural comparison)\
→ Network analysis for peer effects\
→ Expand to more upper-middle income countries
:::

------------------------------------------------------------------------

## Conclusions {.center}

::: {.callout-tip icon="false"}
## Main Finding

**Schools with high concentrations of disadvantaged students produce less learning, even with identical resources**

→ Poverty is not just an individual burden; it creates *systemic disadvantages* through peer effects
:::

:::: fragment
::: incremental
-   Education systems **reproduce** inequality more than they **mitigate** it
-   Non-cognitive skills offer a **malleable, high-return intervention point**
-   Policy must address **both supply** (school quality) **and demand** (student capacity)
:::
::::

------------------------------------------------------------------------

## Thank You {.center}

::: {style="font-size: 1.8em; margin-top: 1em;"}
**Questions?**
:::

::: {style="margin-top: 2em;"}
Garen Avanesian\
Southern Federal University\
📧 avanesian\@sfedu.ru

📊 GitHub: [github.com/karavan88/EduProdFunctionPISA](https://github.com/karavan88/EduProdFunctionPISA)
:::